---
title: 'Data Wrangling Exercise 2: Dealing with missing values'
author: "Louis MS"
date: "10 January 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# library key packages
library(dplyr)
library(tidyr)

# read in CSV file
titanic_original <- read.csv("titanic_original.csv")

```

## Introduction & Method

the data set 'titanic_original.csv' is a data frame with 1310 observations across 15 variables.  In this exercise the data has missing values that require filling. Missing values will be replaced in the following steps.

### Step 1 
Replace all missing values in 'embarked' with 'S'.

```{r}

titanic_st1 <- titanic_original

titanic_st1$embarked[is.na(titanic_st1$embarked)] <- "S"
titanic_st1$embarked <- sub("^$", "S", titanic_st1$embarked)

unique(titanic_st1$embarked)

```


### Step 2
All missing values in 'Age' will be replaced with the mean of all values in 'Age'.

```{r}

titanic_st2 <- titanic_st1

mean(titanic_st2$age, na.rm=TRUE)
# age column mean is 29.88113

median(titanic_st2$age, na.rm=TRUE)
# age column median is 28

titanic_st2$age[is.na(titanic_original$age)] <- mean(titanic_st2$age, na.rm=TRUE)
```

This code has replaced missing age values with the mean of all other age values which came out to be 29.88113.  This decimal number is an unusual number to put in for age, so it maybe more appropriate to replace missing age values with the median value instead as this is an integer (28).

### Step 3
Replace missing values in 'boat' column with 'None' or 'NA' string.

```{r}

titanic_st3 <- titanic_st2
titanic_st3$boat <- sub("^$", "None", titanic_st3$boat)

```

### Step 4
Create a new column 'has_cabin_number' with numerical definition '1' or '0'.

```{r}

titanic_st4 <- titanic_st3
titanic_st4$has_cabin_number <- ifelse(titanic_st4$cabin == "", 0, 1)

```

### Step 5
Read and write cleanded dataset to 'titanic_clean.csv'.

```{r}

titanic_clean <- titanic_st4
write.csv(titanic_clean, file = "titanic_clean.csv")
```

Please note that it is advisable to not create the data frames titanic_st1:4 as these are only slightly varied duplicates of their previous.  I have split them out into their steps for clarity in this exercise.